The importance of a complete description of dependence between variables for ecology and related fields

### 2019-05-02 Paper.Rmd
editing for length

---
title: "The importance of a complete statistical description of dependence between variables for ecology and related fields"
author: "Shyamolina Ghosh, Lawrence W. Sheppard, Mark T. Holder, Terrance E. Loecke, Philip C. Reid, James D. Bever, Daniel C. Reuman"
date: ""
geometry: "left=1.5cm,right=1.5cm,top=2.5cm,bottom=2.8cm"

@@

*Affiliations:*

Ghosh: Department of Ecology and Evolutionary Biology and Kansas Biological Survey, University of Kansas, Lawrence, KS, 66045, USA

Sheppard: Department of Ecology and Evolutionary Biology and Kansas Biological Survey, University of Kansas, Lawrence, KS, 66045, USA

Holder: Department of Ecology and Evolutionary Biology and Biodiversity Institute, University of Kansas, Lawrence, KS, 66045, USA

Loecke: Environmental Studies Program and Kansas Biological Survey, University of Kansas, Lawrence, KS, 66047, USA

Reid: School of Biological & Marine Sciences, University of Plymouth, Drake Circus, Plymouth PL4 8AA, United Kingdom

Bever: Department of Ecology and Evolutionary Biology and Kansas Biological Survey, University of Kansas, Lawrence, KS, 66045, USA

Reuman: Department of Ecology and Evolutionary Biology and Kansas Biological Survey, University of Kansas, Lawrence, KS, 66047, USA; Laboratory of Populations, Rockefeller University, 1230 York Ave, New York, NY, 10065, USA

*Possible alternative titles:*

- A complete statistical description of dependence between variables for ecology and related fields

*Keywords:*

*Type of Article*

- Ideas and Perspectives

*The number of words in the abstract*

*The number of words in the main text (excluding abstract, acknowledgements, references, table and figure legends)*

*The number of words in each text box*

*The number of references*

*The number of figures, tables, and text boxes*

*The name and complete mailing address (including telephone and fax numbers and e-mail address) of the person to whom correspondence should be sent*

\newpage
# Abstract

All branches of ecology study relationships among environmental and biological variables. However, ubiquitously used approaches to studying such relationships, based on correlation and regression, provide only a small slice of the complex information content of the relationships. Other statistical tools exist that provide a complete description of relationships, based on the concept of the *copula*; they are applied in finance, neuroscience and other fields, but not in ecology. We here examine the potential for copula approaches to improve our understanding of ecology. We find that informative copula structure in dependences between variables is exceedingly common across all the environmental, species-trait, phenological, population, community, and ecosystem functioning dataset we considered. Many datasets exhibited asymmetric tail dependence, whereby two variables are more strongly related in their left compared to right tails, or vice versa. Copula structure and tail dependence can help reveal causal relationships between variables. We describe two mechanisms by which observed copula structure and tail dependence can arise in ecological data: a Moran-like effect whereby dependence structures between environmental variables are inherited by ecological variables; and asymmetric or nonlinear influences of environments on ecological variables, such as under Liebig's law of the minimum. We describe consequences of copula structure, including impacts on extinction risk and the stability through time of ecosystem services. By documenting the importance of a complete statistical description of dependence, advancing conceptual frameworks, and demonstrating powerful methods, we aim to elevate copula approaches to widespread use in ecology.

@@

Very generally, all branches of ecology and related areas <!--DAN : Reid said " Is this text needed. 
Ecology covers 
all branches of the biological sciences?
What do you mean by related areas? 
You use related fields later.
Needs clarifying."-->
study relationships among environmental and biological 
variables. However, commonly used correlation and regression 
approaches to studying 
such relationships are limited - such statistical
approaches provide only a very partial description of a relationship between variables. 
For instance, datasets showing markedly different 
relationships may nevertheless have the same correlation coefficient value 
(Fig. \ref{fig_cop_pedag1}A-B). The variables of Fig. \ref{fig_cop_pedag1}A 
(respectively, Fig. \ref{fig_cop_pedag1}B) are principally 
related in the left (respectively, right) portions of their distributions, an asymmetric 
pattern of dependence that may
have ecological significance, as discussed below, but is not captured by 
correlation. Panels C and D 
were obtained from A and B through monotonic transformation 
(Appendix \ref{SM-pedagog1_details} for details),
which alters the Pearson, but not the Spearman correlation. Although correlation
is only one of numerous metrics for 
studying relationships between variables, other common metrics <!--DAN : add example--> also 
provide only partial information.

@@

Other well-developed statistical tools exist, that are 
applied widely in other fields (e.g. finance, neuroscience, bioinformatics etc.), 
and provide a demonstrably complete 
description of the dependence 
between two or more variables [@joe2014_dependence; @nelsen2006_copula]. 
The goal of this paper is to illustrate the potential for 
broader application of these
tools in ecology and related fields. The tools are based on 
the statistical concept of the *copula*, which we here 
introduce. Copula approaches separate the information contained in a bivariate 
random variable into two distinct parts: the information contained in the 
marginal distributions (which says nothing about the dependence between the 
variables), and the remaining information (which is solely about
the dependence between the variables). From a data perspective, given 
a sample $(x_i,y_i)$, $i=1,2,\ldots,n$ (e.g., 
Fig. \ref{fig_cop_pedag1}A-B), the associated dependence
or copula structure is revealed by the bivariate plot of $u_i$ versus $v_i$, where
$u_i$ is the rank of $x_i$ in the set of the $x_j$ ($j=1,\ldots,n$), divided by
$n+1$; and $v_i$ is defined in the same way but using the $y_i$. See, 
for instance,
Fig. \ref{fig_cop_pedag1}E-F, which show the copula structure of 
A-B, respectively. Ranking makes marginals of 
the underlying random variables uniformly distributed, isolating the 
dependence structure only<!--DAN : check this-->.
Mathematically, if $(X,Y)$ is a continuous bivariate random
variable and the cumulative distribution functions (cdfs) of $X$ and $Y$ are 
$F_X(x)=P[X\leq x]$ and $F_Y(y)=P[Y\leq y]$, then the dependence or copula structure
of $(X,Y)$ is revealed by the random variable $(F_X(X),F_Y(Y))$.
This random variable takes values in the unit square, 
$[0,1] \times [0,1]$, and has uniform marginal distributions. Probability density 
functions (pdfs) corresponding to the copulas associated 
with Fig. \ref{fig_cop_pedag1}E-F are in Fig.
\ref{fig_cop_pedag1}G-H, respectively. 
See @nelsen2006_copula and @joe2014_dependence
for general background. A theorem of A. @sklar1959_theorem
states that copulas give a unique and complete 
representation of the dependence
between variables, in the sense that the joint distribution of $(X,Y)$ 
can be reconstructed from the 
copula and the marginal distributions. 
Copula theory is also well developed for
multivariate distributions [@joe1997multivariate].

A principal benefit of a copula approach is 
that it can sensitively detect and model
tail dependence and asymmetries of tail dependence. 
Tail dependence is association between extreme values of variables. 
If smaller values of two positively associated variables are more 
strongly associated than are larger values, the variables are said to
have stronger lower- or left-tail dependence than right- or upper-tail dependence
(Fig. \ref{fig_cop_pedag1}A, E);
and \emph{vice versa} if larger values are more strongly associated than smaller
ones (Fig. \ref{fig_cop_pedag1}B, F). Datasets of the same 
correlation can have a range of tail dependence 
characteristics (Fig. \ref{SM-fig_cop_pedag}), which can be quantified (see section on
methods for Q1).

One way we may expect a copula approach 
to better reflect ecological relationships relates to Liebig's law of the
minimum. If, for instance, the growth of a plant depends on soil 
nitrogen, N, as well
as on other factors, a plot of plant growth rates versus soil nitrogen may be 
expected to look more like Fig. \ref{fig_cop_pedag1}A or E than like 
Fig. \ref{fig_cop_pedag1}B or F. Nitrogen 
will control plant growth, producing a clear
relationship, only when it is limiting, in the left portion of the N and 
plant-growth-rate distributions. 
Nitrogen will be less related or unrelated to growth rates in the right tails of 
the distributions, as in Fig. \ref{fig_cop_pedag1}A and E. The portion
of the distributions that dominate the association is visible in the copula
approach, but not via standard techniques.

A second reason we may expect a copula-based approach to benefit 
ecological research is that such an approach may help illuminate 
the causal mechanisms by which variables are related. For instance, if two species
X and Y are strong competitors, we may expect abundances of the two species
across quadrats to be negatively related, as in 
Fig. \ref{fig_cop_pedag1}I-J. If X is plotted on the horizontal axis
and Y is plotted on the vertical, then the plot in Fig. \ref{fig_cop_pedag1}I 
suggests X is the dominant 
competitor: when X is abundant, Y is necessarily rare because it is suppressed 
by X; whereas when X is rare, Y
may be abundant, or may also be relatively rare due to 
limiting factors other than X. 
Alternatively, Fig. \ref{fig_cop_pedag1}J suggests that Y
is the dominant competitor. The copula approach again reveals which portions of the
distributions are principally responsible for the association between variables, and this 
detail may help support or refute alternative causal hypotheses about 
the relationship between the variables.

A third reason we may expect a copula approach to benefit ecological
research is that prior work has demonstrated important copula 
stucture in the spatial dependence of environmental 
variables [@Serinaldi2008; @She2018; @Goswami2018;
@li2013]. One may expect copula structure in environmental variables 
to take the form of strong tail dependence if
intense meteorological events are also widespread in their effects, as seems 
frequently to be the case.
Spatial dependence in an environmental variable 
tends to beget dependence across space
in fluctuations of populations or other ecosystem variables 
(this is called \emph{spatial synchrony}) which are influenced by the environmental 
variable. This process is called the \emph{Moran effect}. 
If copula structure is transmitted, 
possibly in modified form, 
from environmental to ecological variables, then we may expect important 
copula structure to be a common feature of spatial synchrony of population, community, 
biogeochemical and other ecological variables.
Synchrony is a phenomenon of great interest in 
ecology [@Liebhold2004; @Sheppard2016; @walteretal2017].

A fourth reason we may expect copula structure to be important
in ecology has to do with spatially aggregated or averaged quantities. 
Many ecological variables of applied importance 
depend on the spatial average or sum of local 
quantities. For instance, regional methane and CO$_2$ 
fluxes are the sum of local fluxes; 
and the total economic value of a fishery is the sum of catches from 
local areas. We will explore how details, captured by copula methods, of the 
nature of the dependence between local quantities can influence the spatial 
mean or sum, and how this may influence human concerns and 
higher organizational levels in 
ecology. Illustrating the idea here, prior work demonstrated that the overall 
reliability of wind-generated electricity depends sensitively on details
of the dependence between wind speeds at multiple generator sites [@li2013]. 
Spatially aggregated ecological variables may show a similar dependence, with
potentially substantial consequences. For instance, if populations of a 
pest species in different locations are all positively associated and also strongly
related to each other in their right tails, then local outbreaks tend 
to occur together, creating regional epidemics. If populations
of a crop or other resource species are strongly 
related in their left tails, then 
episodes of local scarcity occur together, creating regional 
shortages. On the other hand left-tail dependence in the pest species 
or right-tail dependence in the crop, even if overall correlation were 
unchanged, would have much more benign effects.

We approached our overall goal of exploring the potential of 
copula approaches in ecology by addressing the 
following specific questions. (Q1) Do datasets in ecology and related fields 
have copula structure distinct from that of a Gaussian/normal distribution (here called 
non-normal copula structure)? Do they show tail dependence distinct from that 
of a normal distribution, and in particular do they show asymmetric tail 
dependence? (Q2) If so, what are some possible causes/mechanisms of this 
non-normal copula structure? And (Q3) what are the consequences of non-normal 
copula structure and tail dependence for our understanding of biology and for 
applications? To address Q1, we analyze several datasets, including 
environmental, species-trait, phenological, population, community, and 
ecosystem functioning data. We selected these datasets to span multiple 
sub-fields and organizational levels of ecology. The second section of the 
paper introduces the data, and the third and fourth sections use the data to 
address Q1. The fifth and sixth sections, which address Q2, use simple models to 
explore mechanisms. The seventh and eigth sections, which address Q3, use both
the data of section two as well as models to explore a variety of consequences
of copula structure and tail dependence for applications and for understanding
ecology. The ninth section is the Discussion. 
Copula approaches have been used to great benefit in neuroscience [@onken2009], 
bioinformatics 
[@Kim2008], medical research [@Emura2016], direct study of environmental 
variables [@Serinaldi2008; @She2018; @Goswami2018; @li2013], finance, and other fields. 
To our knowledge, these approaches have never or seldom been used for ecological research. 
We argue that benefits to ecology of using a complete statistical 
desription of dependence will be substantial.

@@

Datasets included environmental, species-trait, phenological, 
population, community, and ecosystem functioning data (Table \ref{tab_data_info}). Copula structure describing 
the relationship between atmospheric weather variables such as rainfall or wind speed 
as measured in multiple locations has been 
examined previously in the meteorological literature [e.g., @Serinaldi2008; @li2013]. We therefore 
examined environmental variables from the soil instead, using the Rapid Carbon 
Assessment database [RaCA; @Wills2014].
<!--***DAN: Add Wills and Loecke pending reference if published in time--> 
The database comprises measurements of soil organic carbon and total soil 
nitrogen (Mg C or N per hectare of soil surface) from `r nrow(d_soilCN)` 
locations across the coterminous United States (Fig. \ref{SM-fig_SoilCNmapUSA} for sampling locations).
Measurments were from the central pedon of a soil core, to 1 m depth.
Species-trait data were average species basal metabolic rate (BMR, KJ per hour) and 
body mass (grams) data for `r nrow(d_birdsBMR)`  species of birds [@Mcnab_2009] 
and `r nrow(d_mammalsBMR)`  species of mammals [@Mcnab_2008].
These data have been much studied, but to our knowledge the copula structure 
of the dependence between metabolic rate and body mass has not been examined. 
Species-trait data such as these reflect the coevolution of the two traits.
Phenological data were aphid first-flight dates from 10 locations (Fig. \ref{SM-fig_aphid_map}) 
across the United Kingdom (UK) for 20 aphid species (Table \ref{SM-tab_aphid_info}), for the 
35 years 1976 to 2010. These time series were computed from the Rothamsted Insect 
Survey (RIS) suction-trap dataset [@Harrington2014; @Bell2015]. The first of our two 
population-level datasets was also derived from the RIS suction-trap data, and comprised 
total counts of aphids trapped for the same locations, species, and years; these are 
indicators of aphid migration counts. Data pre-processing for the aphid data was the same as 
that of @Sheppard2016,
but we excluded one of the 11 sites they used in order to only use sites with at least 30 
years data.  Our second population dataset comprised average annual plankton abundance estimates 
for 14 locations (Fig. \ref{SM-fig_plankton_map})
in the North Sea and British seas for 22 taxa (Table \ref{SM-tab_plankton_info}) for the 
56 years 1958 to 2013. The locations are $2^\circ$ by $2^\circ$ grid cells. These data 
were obtained from the Continuous Plankton Recorder (CPR) dataset, now operated and maintained by 
the Marine Biological Association of the United Kingdom. Data pre-processing was similar but not identical to that 
of @Sheppard2017; see Appendix \ref{SM-Data_details} for details.
Community-level data, obtained from the Cedar Creek Ecosystem Science Reserve, were 
plant aboveground biomass and Shannon's diversity index for `r nrow(d_cc[[5]])` plots, each 9m by 9m, as sampled in 
the years 1996-2000 and 2007, each year analyzed separately [@Tilman2001; @Tilman2006]. 
Shannon's index was computed from plant species percent cover data.  
Finally, ecosystem functioning data were methane (CH$_{4}$) fluxes between the soil or 
water surface and the troposphere, measured at 13 locations at daily to weekly intervals 
from September 2015 to September 2016 at the Great Miami Wetland mitigation bank in Trotwood, 
Ohio [@Holland1999; @Jarecke2016; @smyth2019using]. Each included location was measured on at least 50 dates. 
Sampling locations spanned a hydrologic gradient from upland to seasonal wetland.
See Appendix \ref{SM-Data_details} for additional data details.

The environmental, species-trait, and community datasets are henceforth 
called the "bivariate" datasets in the sense that they comprise two quantities measured at 
different locations or for different species. The phenology and population datasets are 
"multivariate" in that they comprise, for each taxon, measurements through time at multiple locations; 
copula structure can be studied for each pair of locations. Likewise the ecosystem 
functioning (methane flux) dataset was  multivariate in this sense (Table \ref{tab_data_info}).

@@

## Results for Q3: Tail dependence influences skewness of spatial average, extinction risk, and Taylor's law

>>>>

---
title: "The importance of a complete description of dependence between variables for ecology and related fields"
author: "Shyamolina Ghosh, Lawrence W. Sheppard, Mark T. Holder, Terrance E. Loecke, Philip C. Reid, James D. Bever, Daniel C. Reuman"
date: ""
geometry: "left=1.5cm,right=1.5cm,top=2.5cm,bottom=2.8cm"

@@

\noindent \emph{Affiliations:}

\noindent Ghosh: Department of Ecology and Evolutionary Biology and Kansas Biological Survey, University of Kansas, Lawrence, KS, 66045, USA

\noindent Sheppard: Department of Ecology and Evolutionary Biology and Kansas Biological Survey, University of Kansas, Lawrence, KS, 66045, USA

\noindent Holder: Department of Ecology and Evolutionary Biology and Biodiversity Institute, University of Kansas, Lawrence, KS, 66045, USA

\noindent Loecke: Environmental Studies Program and Kansas Biological Survey, University of Kansas, Lawrence, KS, 66047, USA

\noindent Reid: School of Biological & Marine Sciences, University of Plymouth, Drake Circus, Plymouth PL4 8AA, United Kingdom

\noindent Bever: Department of Ecology and Evolutionary Biology and Kansas Biological Survey, University of Kansas, Lawrence, KS, 66045, USA

\noindent Reuman: Department of Ecology and Evolutionary Biology and Kansas Biological Survey, University of Kansas, Lawrence, KS, 66047, USA; Laboratory of Populations, Rockefeller University, 1230 York Ave, New York, NY, 10065, USA

\noindent \emph{Possible alternative titles:}

\noindent A complete statistical description of dependence between variables for ecology and related fields

\noindent \emph{Keywords:}

\noindent dependence, regression, correlation, copula, population, community

\noindent \emph{Type of Article:}

\noindent Ideas and Perspectives

\noindent \emph{The number of words in the abstract:}

\noindent 241

\noindent \emph{The number of words in the main text (excluding abstract, acknowledgements, references, table and figure legends):}

\noindent \emph{The number of words in each text box:}

\noindent No text boxes

\noindent \emph{The number of references:}

\noindent \emph{The number of figures, tables, and text boxes:}

\noindent 7 figures, 3 tables, 0 text boxes

\noindent \emph{The name and complete mailing address (including telephone and fax numbers and e-mail address) of the person to whom correspondence should be sent:}

\noindent Daniel Reuman, 2101 Constant Ave, Lawrence, KS, 66047, reuman@ku.edu, 626 560 7084.

\newpage

# Abstract

All branches of ecology study relationships among environmental and biological variables. However, ubiquitously used approaches to studying such relationships, based on correlation and regression, provide only a small slice of the complex information contained in the relationships. Other statistical tools exist that provide a complete description of relationships between variables, based on the concept of the *copula*; they are applied in finance, neuroscience and other fields, but not in ecology. We here explore the concepts that underpin copulas and examine the potential for those concepts to improve our understanding of ecology. We find that informative copula structure in dependences between variables is exceedingly common across all the environmental, species-trait, phenological, population, community, and ecosystem functioning dataset we considered. Many datasets exhibited asymmetric tail dependence, whereby two variables are more strongly related in their left compared to right tails, or vice versa. Copula structure and tail dependence can help reveal causal relationships between variables. We describe mechanisms by which observed copula structure and tail dependence can arise in ecological data, including a Moran-like effect whereby dependence structures between environmental variables are inherited by ecological variables; and asymmetric or nonlinear influences of environments on ecological variables, such as under Liebig's law of the minimum. We describe consequences of copula structure, including impacts on extinction risk and the stability through time of ecosystem services. By documenting the importance of a complete description of dependence, advancing conceptual frameworks, and demonstrating a powerful approach, we aim to elevate copula reasoning to widespread use in ecology.

@@

Very generally, all branches of ecology  
study relationships among environmental and biological 
variables. However, commonly used correlation and regression 
approaches to studying 
such relationships are limited - such 
approaches provide only a very partial description of a relationship between variables. 
For instance, datasets showing markedly different 
relationships may nevertheless have the same correlation coefficient value 
(Fig. \ref{fig_cop_pedag1}A-B). The variables of Fig. \ref{fig_cop_pedag1}A 
(respectively, Fig. \ref{fig_cop_pedag1}B) are principally 
related in the left (respectively, right) portions of their distributions, an asymmetric 
pattern of dependence that may
have ecological significance, as discussed below, but is not captured by 
correlation. Panels C and D 
were obtained from A and B through monotonic transformation 
(Appendix \ref{SM-pedagog1_details} for details),
which alters the Pearson, but not the Spearman correlation. Although correlation
is only one way to 
study relationships between variables, other common metrics also 
provide partial information.

@@

Well-developed approaches exist, and are 
applied widely in finance, neuroscience, bioinformatics and other fields, 
that provide a description of the dependence 
between two or more variables that can be mathematically proven 
to be complete [@joe2014_dependence; @nelsen2006_copula]. 
The goal of this paper is to illustrate the potential for 
broader application of the concepts behind these approaches 
in ecology. The approaches are based on 
the statistical concept of the *copula*, which we here 
introduce. Copula approaches separate the information contained in a bivariate 
random variable into two distinct parts: the information contained in the 
marginal distributions (which says nothing about the dependence between the 
variables), and the remaining information (which is solely about
the dependence). From a data perspective, given 
a sample $(x_i,y_i)$, $i=1,2,\ldots,n$ (e.g., 
Fig. \ref{fig_cop_pedag1}A-B), the associated dependence
structure is revealed by the bivariate plot of $u_i$ versus $v_i$, where
$u_i$ is the rank of $x_i$ in the set of the $x_j$ ($j=1,\ldots,n$), divided by
$n+1$; and $v_i$ is defined in the same way but using the $y_i$. See, 
for instance,
Fig. \ref{fig_cop_pedag1}E-F, which show the copula structure of 
A-B, respectively. Ranking makes marginals of 
the underlying random variables uniformly distributed, isolating the 
dependence structure only.
Mathematically, if $(X,Y)$ is a continuous random
variable and the cumulative distribution functions (cdfs) of $X$ and $Y$ are 
$F_X(x)=P[X\leq x]$ and $F_Y(y)=P[Y\leq y]$, then the dependence structure
of $(X,Y)$ is revealed by the random variable $(F_X(X),F_Y(Y))$.
This random variable takes values in $[0,1] \times [0,1]$, and has 
uniform marginal distributions. Probability density 
functions (pdfs) corresponding to the copulas associated 
with Fig. \ref{fig_cop_pedag1}E-F are in Fig.
\ref{fig_cop_pedag1}G-H, respectively. 
See @nelsen2006_copula and @joe2014_dependence
for general background. A theorem of @sklar1959_theorem
states that copulas give a unique and complete 
representation of the dependence
between variables.

A principal benefit of a copula approach is 
that it can sensitively detect tail dependence and asymmetries of tail dependence. 
Tail dependence is association between extreme values of variables. 
If smaller values of two positively associated variables are more 
strongly associated than are larger values, the variables are said to
have stronger lower- or left-tail dependence than right- or upper-tail dependence
(Fig. \ref{fig_cop_pedag1}A, E);
and \emph{vice versa} if larger values are more strongly associated than smaller
ones (Fig. \ref{fig_cop_pedag1}B, F). Datasets of the same 
correlation can have a range of tail dependence 
characteristics (Fig. \ref{SM-fig_cop_pedag}), which can be quantified (see section on
methods for Q1).

One way we may expect a copula approach 
to better reflect ecological relationships relates to Liebig's law of the
minimum. If, for instance, the growth of a plant depends on soil 
nitrogen, N, and other factors, a plot of plant growth rates versus N may 
look more like Fig. \ref{fig_cop_pedag1}A or E than like 
Fig. \ref{fig_cop_pedag1}B or F. Nitrogen 
will control plant growth, producing a clear
relationship, only when it is limiting, in the left portions of the distributions. 
The portion of the distributions that dominate the association is visible in the copula
approach, but not via standard techniques.

A second reason we may expect a copula approach to benefit 
ecological research is that such an approach may help illuminate 
the causal mechanisms between variables. For instance, if two species
X and Y are strong competitors, abundances of the two species
across quadrats should be negatively related, as in 
Fig. \ref{fig_cop_pedag1}I-J. If X is plotted on the horizontal axis
and Y is on the vertical, then Fig. \ref{fig_cop_pedag1}I 
would suggest X is the dominant 
competitor: when X is abundant, Y is necessarily rare because it is suppressed; 
whereas when X is rare, Y
may be abundant, or may also be relatively rare due to 
limiting factors other than X. 
Alternatively, Fig. \ref{fig_cop_pedag1}J suggests that Y
is the dominant competitor. The copula approach again reveals which portions of the
distributions are principally responsible for the association between variables, and this 
information can help support or refute alternative causal hypotheses.

A third reason we may expect a copula approach to benefit ecological
research is that prior work has demonstrated important copula 
stucture in the spatial dependence of environmental 
variables [@Serinaldi2008; @She2018; @Goswami2018;
@li2013]. One may expect copula structure in environmental variables 
to take the form of strong tail dependence if
intense meteorological events are also widespread in their effects, as seems 
frequently to be the case.
Spatial dependence in an environmental variable 
tends to beget spatial dependence 
in fluctuations of populations or other ecosystem variables 
(this is called \emph{spatial synchrony}) which are influenced by the environmental 
variable. This is called the \emph{Moran effect}. 
If copula structure is transmitted, 
possibly in modified form, 
from environmental to ecological variables, then we may expect important 
copula structure to be a common feature of the spatial synchrony of population, community, 
biogeochemical and other ecological variables.
Synchrony is a phenomenon of great interest in 
ecology [@Liebhold2004; @Sheppard2016; @walteretal2017].

A fourth reason we may expect copula structure to be important
in ecology has to do with spatially aggregated or averaged quantities. 
Many ecological variables of applied importance 
depend on the spatial average or sum of local 
quantities. For instance, regional methane and CO$_2$ 
fluxes are the sum of local fluxes; 
and the total economic value of a fishery is the sum of local catches. 
We will explore how details, captured by copulas, of the 
dependence between local quantities can influence the spatial 
mean or sum, and how this may influence higher organizational levels in 
ecology and human concerns. Illustrating the idea here, prior work demonstrated that the overall 
reliability of wind-generated electricity depends sensitively on details
of the dependence between wind speeds at multiple generator sites [@li2013]. 
Spatially aggregated ecological variables may show a similar dependence, with
potentially substantial consequences. For instance, if populations of a 
pest species in different locations are all positively associated and also strongly
related to each other in their right tails, then local outbreaks will tend 
to occur together, creating regional epidemics. 
On the other hand left-tail dependence in a pest species, even if overall 
correlation were unchanged, would have much more benign effects.

We approached our overall goal of exploring the potential of 
copula approaches for ecology by addressing the 
following specific questions. (Q1) Do datasets in ecology  
have copula structure distinct from that of a Gaussian/normal distribution (here called 
non-normal copula structure)? Do they show tail dependence distinct from that 
of a normal distribution, and in particular do they show asymmetric tail 
dependence? (Q2) If so, what are some possible causes/mechanisms of this 
non-normal copula structure? And (Q3) what are the consequences of non-normal 
copula structure and tail dependence for our understanding of ecology and for 
applications? To address Q1, we analyze several datasets, including 
environmental, species-trait, phenological, population, community, and 
ecosystem functioning data. We selected these datasets to span multiple 
sub-fields and organizational levels of ecology. The second section of the 
paper introduces the data, and the third and fourth sections use the data to 
address Q1. The fifth and sixth sections, which address Q2, use simple models to 
explore mechanisms. The seventh and eigth sections, which address Q3, use both
data and models to explore a variety of consequences. The ninth section is the Discussion. 
Copula approaches have been used to great consequence in neuroscience [@onken2009], 
bioinformatics 
[@Kim2008], medical research [@Emura2016], direct study of environmental 
variables [@Serinaldi2008; @She2018; @Goswami2018; @li2013], and finance [@Li2000]. 
To our knowledge, these concepts have never been used in ecology. 
We argue that benefits to ecology of using a complete 
desription of dependence will be substantial.

@@

Datasets included environmental, species-trait, phenological, 
population, community, and ecosystem functioning data (Table \ref{tab_data_info}). Copula structure describing 
the relationship between atmospheric weather variables such as rainfall or wind speed 
as measured in multiple locations has been 
examined previously in the meteorological literature [e.g., @Serinaldi2008; @li2013]. We therefore 
examined environmental variables from the soil instead, using the Rapid Carbon 
Assessment database [RaCA; @Wills2014].
<!--***DAN: Add Wills and Loecke pending reference if published in time--> 
The database comprises measurements of soil organic carbon and total soil 
nitrogen (Mg C or N per hectare of soil surface) from `r nrow(d_soilCN)` 
locations across the coterminous United States (Fig. \ref{SM-fig_SoilCNmapUSA} for sampling locations).
Species-trait data were average species basal metabolic rate (BMR, KJ per hour) and 
body mass (grams) for `r nrow(d_birdsBMR)`  species of birds [@Mcnab_2009] 
and `r nrow(d_mammalsBMR)`  species of mammals [@Mcnab_2008].
These data have been much studied, but to our knowledge the copula structure 
of the dependence between metabolic rate and body mass has not been examined. 
Species-trait data such as these reflect the coevolution of the two traits.
Phenological data were aphid first-flight dates from 10 locations (Fig. \ref{SM-fig_aphid_map}) 
across the United Kingdom (UK) for 20 aphid species (Table \ref{SM-tab_aphid_info}), for the 
35 years 1976 to 2010. These time series were computed from the Rothamsted Insect 
Survey (RIS) suction-trap dataset [@Harrington2014; @Bell2015]. The first of our two 
population-level datasets was also derived from the RIS suction-trap data, and comprised 
total counts of aphids trapped for the same locations, species, and years. 
Our second population dataset comprised average annual plankton abundance estimates 
for 14 locations (Fig. \ref{SM-fig_plankton_map})
in the North Sea and British seas for 22 taxa (Table \ref{SM-tab_plankton_info}) for the 
56 years 1958 to 2013. The locations are $2^\circ$ by $2^\circ$ grid cells. 
Community-level data, obtained from the Cedar Creek Ecosystem Science Reserve, were 
plant aboveground biomass and Shannon's diversity index for `r nrow(d_cc[[5]])` plots, each 9m by 9m, as sampled in 
the years 1996-2000 and 2007, each year analyzed separately [@Tilman2001; @Tilman2006]. 
Shannon's index was computed from plant species percent cover data.  
Finally, ecosystem functioning data were methane (CH$_{4}$) fluxes between the soil or 
water surface and the troposphere, measured at 13 locations at daily to weekly intervals 
from September 2015 to September 2016 at the Great Miami Wetland mitigation bank in Trotwood, 
Ohio [@Holland1999; @Jarecke2016; @smyth2019using]. Each included location was measured on at least 50 dates. See Appendix \ref{SM-Data_details} for additional data details.

The environmental, species-trait, and community datasets are henceforth 
called the "bivariate" datasets in the sense that they comprise two quantities measured at 
different locations or for different species (Table \ref{tab_data_info}). 
The phenology, population, and ecosystem functioning datasets are 
"multivariate" in that they comprise, for each taxon, measurements through time at multiple locations; 
copula structure can be studied for each pair of locations.

<!--***DAN: made it to here in my editing for length, 2019 05 02-->

@@

# Results for Q3: Tail dependence influences skewness of spatial average, extinction risk, and Taylor's law
